Finally Tagless, Partially Evaluated

### 2007-07-27 tagless/tagless-final.tex
Small formatting changes to the JFP version that Jacques suggested: - for Figure 2, since we now have space. might it make sense to separate it into 2 figures (the embedding and the example)?  It certainly would make the figure clearer and simplify the text. - JFP discourages footnotes -- should we eliminate them all?

\begin{figure*}[t]
\begin{floatrule}
\begin{code2}
module type Symantics = sig type ('c, 'dv) repr
  val int : int  -> ('c, int) repr
  val bool: bool -> ('c, bool) repr

@@

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c, bool) repr
            -> (unit -> 'x) -> (unit -> 'x) -> (('c, 'da) repr as 'x)
end

module EX(S: Symantics) = struct open S
  let test1 () = app (lam (fun x -> x)) (bool true)
  let testpowfix () =
       lam (fun x -> fix (fun self -> lam (fun n ->
        if_ (leq n (int 0)) (fun () -> int 1)
            (fun () -> mul x (app self (add n (int (-1))))))))
  let testpowfix7 = lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code2}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language, 
    and examples of use}
\label{fig:ocaml-simple}
\end{figure*}
To embed the same object language in (Meta)OCaml, we replace the \ifshort type
class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |'c|, an
\emph{environment classifier} \cite{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

The functor |EX| in Fig.~\ref{fig:ocaml-simple} encodes 
our running examples |test1| and the $\mathit{power}$ function
(|testpowfix|).
The dummy argument to |test1| and |testpowfix| is an artifact of
MetaOCaml, related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction dictates that
the definitions of our object terms must look syntactically like
values. Alternatively, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

(d) \BUseVerbatim{5d}
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
\fi
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1| with
\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
\ifshort\pagebreak[3]\fi
with
\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
and does not
call the interpreter, unlike the recursive calls to |eval0| and
|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
The resulting code obviously contains no tags and no pattern matching.
\ifshort
The accompanying code shows the Haskell implementation. 
\else
\begin{proposition}
    The compiler |C| generates tagless, pattern-match free code.
\end{proposition}
This is syntactically clear from the body of |C|, and follows
readily by structural induction and the properties of |.~| and
|.< >.|.

We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it: we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.%
\footnote{\texttt{ByteCode} can be mapped to the abstract syntax of Template
    Haskell. The output of our compiler will then be assuredly type-correct
    Template Haskell.}
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. The
accompanying code gives the full details.
\fi

@@

\fi
Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(of type |C Bool|), the second a static term (of type |R a|) and the third a
dynamic term (of type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.\ifshort\else
\footnote{If we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
f = app (lift f)}}\fi

@@

\fi
This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml\ifshort\else\ as GADTs are unavailable there\fi.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in
|app|\ifshort. \else\
above: the type |P t| has four constructors, of which the pattern in
|app| matches only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
\fi
\ifshort
Although Coq or logical
frameworks with canonical forms could prove pattern\hyp matching
exhaustive,
\else
Indeed the metalanguage implementation could reason thus,%
\footnote{If we use inductive families (as those in Coq) or logical
  frameworks with canonical forms (as in Twelf with its coverage checker),
  we can prove the pattern matching to be exhaustive.}
but it may not;
\fi
GHC cannot and issues warnings.
\ifshort
Recall, our
\else
Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but never fail in reality. The
\fi
goal is to make this exhaustiveness \emph{syntactically} apparent.

@@

Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\ifshort \BUseVerbatim{P} \else \UseVerbatim{P} \fi
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\ifshort\vspace*{-0.8em}
\begin{code3}
repr_pe Int = Int; repr_pe Bool = Bool; repr_pe (a -> b) = P a -> P b
\end{code3}
\vspace*{-0.8em}
\else
\begin{code}
repr_pe Int      = Int
repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code}
\fi
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate
this\ifshort\else\ portability\fi,
we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
% aplas: abbreviated figure
\begin{figure*}
\begin{floatrule}
\ifshort
\begin{code2}
module type Symantics = sig type ('c,'sv,'dv) repr
  val int : int -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
            -> ('c,'x,'da -> 'db) repr
  val app : ('c,'x,'da -> 'db) repr
            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
                                'da -> 'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr
            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
\end{code2}
\else
\begin{code2}
module type Symantics = sig
  type ('c,'sv,'dv) repr

@@

  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr
            -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code2}
\fi
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation \ifshort \protect\linebreak[2] (\texttt{bool},
  \texttt{mul}, \texttt{leq} are elided)\fi}
\label{fig:ocaml}
\end{figure*}

>>>>

\begin{SaveVerbatim}{ocaml-simple}
module type Symantics = sig type ('c, 'dv) repr
  val int : int  -> ('c, int) repr
  val bool: bool -> ('c, bool) repr

@@

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c, bool) repr
            -> (unit -> 'x) -> (unit -> 'x) -> (('c, 'da) repr as 'x)
end
\end{SaveVerbatim}
\begin{SaveVerbatim}{ocaml-examples}
module EX(S: Symantics) = struct open S
  let test1 () = app (lam (fun x -> x)) (bool true)
  let testpowfix () =
       lam (fun x -> fix (fun self -> lam (fun n ->
        if_ (leq n (int 0)) (fun () -> int 1)
            (fun () -> mul x (app self (add n (int (-1))))))))
  let testpowfix7 = lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{SaveVerbatim}

\ifshort
\begin{figure}[t]
\begin{floatrule}
\begin{tabular}{@{}l@{}}
\ifx\relax\normalbaselineskip\else\baselineskip\normalbaselineskip\fi
\BUseVerbatim[baseline=b]{ocaml-simple}\\[\medskipamount]
\ifx\relax\normalbaselineskip\else\baselineskip\normalbaselineskip\fi
\BUseVerbatim[baseline=t]{ocaml-examples}
\end{tabular}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language, 
    and examples of use}
\label{fig:ocaml-simple}
\label{fig:ocaml-examples}
\end{figure}
\else
\begin{figure}[t]
\begin{floatrule}
\BUseVerbatim{ocaml-simple}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure}
\begin{figure}[t]
\begin{floatrule}
\BUseVerbatim{ocaml-examples}
\end{floatrule}
\caption{Examples using the embedding in Figure~\ref{fig:ocaml-simple} of our object language}
\label{fig:ocaml-examples}
\end{figure}
\fi

To embed the same object language in (Meta)OCaml, we replace the \ifshort type
class \fi |Symantics| \ifshort\else type class \fi and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices until~\S\ref{PE}.  The two differences are:
the additional type parameter |'c|, an
\emph{environment classifier} \cite{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| since OCaml is a call-by-value
language.

The functor |EX| in Fig.~\ref{fig:ocaml-examples} encodes 
our running examples |test1| and the $\mathit{power}$ function
(|testpowfix|).
The dummy argument to |test1| and |testpowfix| is an artifact of
MetaOCaml, related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction dictates that
the definitions of our object terms must look syntactically like
values. Alternatively, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

(d) \BUseVerbatim{5d}
\caption{The tagless staged interpreter \texttt{C}.}
\label{fig:interpreter-C}
\end{figure}
\fi
This is a straightforward staging of
|module R|.
This compiler produces
unoptimized code. For example, interpreting our |test1| with
\ifshort Fig.~\ref{fig:interpreter-C}(b) \else \UseVerbatim{5b}\fi
gives the code value |.<(fun|~|x_6|~|->|~|x_6)| |true>.|
of inferred type |('c,|~|bool)| |C.repr|.  Interpreting |testpowfix7|
\ifshort\pagebreak[3]\fi
with
\ifshort Fig.~\ref{fig:interpreter-C}(c) \else \UseVerbatim{5c}\fi
gives a code value with many apparent $\beta$- and $\eta$-redexes\ifshort,
Fig.~\ref{fig:interpreter-C}(d). \else: \UseVerbatim{5d}\fi
This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun|~|x_6|~|->|~|x_6|\ifshort\else\
and does not
call the interpreter, unlike the recursive calls to |eval0| and
|eval| in the |L e| lines in \S\ref{tagproblem}\fi.
The resulting code obviously contains no tags and no pattern matching.
\ifshort
The accompanying code shows the Haskell implementation. 
\else
\begin{proposition}
    The compiler |C| generates tagless code free of pattern matching.
\end{proposition}
This is syntactically clear from the body of |C|, and follows
readily by structural induction and the properties of |.~| and
|.< >.|.

We have also implemented this compiler in Haskell. 
Since Haskell
has no (convenient, typed) staging facility, we had to emulate
it: we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
(Alternatively, we could use Template Haskell as our staging facility:
\texttt{ByteCode} can be mapped to the abstract syntax of Template
Haskell. The output of our compiler will then be assuredly type-correct
Template Haskell.)
Whereas our representation of object terms uses HOAS,
our bytecode uses integer-named
variables to be realistic. 
We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|.
The implementation is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code above. (The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \texttt{ByteCode} \emph{data type}.) The
accompanying code gives the full details.
\fi

@@

\fi
Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(of type |C Bool|), the second a static term (of type |R a|) and the third a
dynamic term (of type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\cite{xi-guarded,WalidPOPL03}.\ifshort\else
(By the way, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\texttt{ fmap
f = app (lift f)~}.)\fi

@@

\fi
This solution using GADTs works but is not quite satisfactory. First, it
cannot be ported to MetaOCaml\ifshort\else, as GADTs are unavailable there\fi.  Second,
the problem of nonexhaustive pattern\hyp matching reappears in
|app|\ifshort. \else\
above: the type |P t| has four constructors, of which the pattern in
|app| matches only |VF| and~|E|. One may say that the
constructors |VI| and |VB| obviously cannot occur because they do not
construct values of type |P (a -> b)| as required by the type of |app|.
\fi
\ifshort
Although Coq or logical
frameworks with canonical forms could prove pattern\hyp matching
exhaustive,
\else
Indeed, the metalanguage implementation could reason thus:
if we use inductive families (as in Coq) or logical
  frameworks with canonical forms (as in Twelf with its coverage checker),
  we can prove the pattern matching to be exhaustive.
Then again, the metalanguage implementation may not reason thus:
\fi
GHC cannot and issues warnings.
\ifshort
Recall, our
\else
Although this point may seem minor, it is the heart of
the tagging problem and the purpose of tag elimination. A typed tagged
interpreter contains many pattern\hyp matching forms that look partial
but never fail in reality. The
\fi
goal is to make this exhaustiveness \emph{syntactically} apparent.

@@

Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\ifshort \BUseVerbatim{P} \else \UseVerbatim{P} \fi
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\ifshort\vspace*{-0.8em}
\begin{code3}
repr_pe Int = Int; repr_pe Bool = Bool; repr_pe (a -> b) = P a -> P b
\end{code3}
\vspace*{-0.8em}
\else
\begin{code}
repr_pe Int      = Int
repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code}
\fi
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate
this\ifshort\else\ portability\fi,
we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
% aplas: abbreviated figure
\begin{figure}
\begin{floatrule}
\ifshort
\begin{code2}
module type Symantics = sig type ('c,'sv,'dv) repr
  val int : int -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
            -> ('c,'x,'da -> 'db) repr
  val app : ('c,'x,'da -> 'db) repr
            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
                                'da -> 'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr
            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
\end{code2}
\else
\begin{code2}
module type Symantics = sig
  type ('c,'sv,'dv) repr

@@

  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val mul : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val leq : ('c,int,int) repr -> ('c,int,int) repr -> ('c,bool,bool) repr
  val if_ : ('c,bool,bool) repr
            -> (unit -> 'x) -> (unit -> 'x) -> (('c,'sa,'da) repr as 'x)
end
\end{code2}
\fi
\end{floatrule}
\caption{A (Meta)OCaml embedding of our object language that supports
  partial evaluation \ifshort \protect\linebreak[2] (\texttt{bool},
  \texttt{mul}, \texttt{leq} are elided)\fi}
\label{fig:ocaml}
\end{figure}
